class: inverse, center, middle
# Descriptive statistics

---
# Summarizing continuous variables

.large[
Functions to summarize continuous variables:
  
  - `summary()`: calculates the minimum, 1st quartile, median, mean, 3rd quartile, and maximum as well as the number of missing values. 
- `mean()`: calculates the mean
- `sd()`: calculates the standard deviation
- `median()`: calculates the median
- `min()`: calculates the minimum
- `max()`: calculates the maximum
- `quantile()`: calculates specified quantiles

Note that in the presence of missing data, most of these functions will require the use of the `na.rm = TRUE` argument, which simply ignores any missing values in the calculation. If not specified and missing values are present, the function will return NA.
]

---
# Some examples using the "marker" variable

```{r}
summary(trial$marker)
mean(trial$marker, na.rm = TRUE)
quantile(trial$marker, probs = c(0.25, 0.75), na.rm = TRUE)
```

Notes:
- Specified the argument `probs = c(0.25, 0.75)` to the `quantile()` function to specify that we wanted the 1st (i.e. 25%) and 3rd (i.e. 75%) quantiles.
- Must use the syntax `dataset$variable` 


---
# Continuous summary by a categorical variable 

.large[We can get summaries of continuous variables by a categorical variable using the function `tapply()`.

For example, summarize "marker" by levels of "grade":]
  
```{r}
tapply(trial$marker, trial$grade, summary)
```

---
# Summarizing categorical variables

.large[
Functions to summarize categorical variables:
  
- `table()`: calculates the number in each group 
- `prop.table()`: calculates the proportion in each group

Note that the argument `useNA = "ifany"` needs to be specified in `table()` in order to see the number of any missing values. If not specified, missing values will be ignored.
]

---
# Example output using the "response" variable

```{r}
table(trial$response, useNA = "ifany")
prop.table(table(trial$response))
```


---
# Cross-tabulations

.large[
Generate cross-tabulations of two categorical variables by passing a second variable to `table()`. For example if we wanted a table of "response" according to "grade":
]

```{r}
table(trial$grade, trial$response, useNA = "ifany")
prop.table(table(trial$grade, trial$response))
```

---
# Summarizing many variables

.large[
What we often need to do in our area of work is what we call a "Table 1" where we describe many characteristics of a study population either overall, or according to another variable. 
There are many packages to do this in R; I will introduce you to the {gtsummary} package.

We installed and loaded this package earlier in order to access the "trial" dataset to use as an example, but as a reminder:
]

```{r eval = FALSE}
install.packages("gtsummary")
library(gtsummary)
```

The summary statistics can be customized in many ways. See [the package website](https://www.danieldsjoberg.com/gtsummary/index.html) for more details and many tutorials.

---
# Overall summary tables

.large[
We can make overall tables using the `tbl_summary()` function. First we will limit the variables in the data to a few of interest, and create a new subset dataset called "trial2":
]
  
```{r}
trial2 <- trial[, c("marker", "grade", "trt")]

tbl_summary(trial2)
```

---
# Cross-tabulated summary tables

.large[
We can also easily create tables of summary statistics according to another variable by specifying the `by` argument to the `tbl_summary()` function:
]

```{r}
tbl_summary(
  trial2, 
  by = "trt"
)
```

.large[
***Note that the functions from the {gtsummary} package are particularly useful when used in conjunction with R Markdown, since they are meant to produce publication-ready papers.***
]